\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Local Adaptive Optimization of Time Step}

### 2013-10-29 report/article.tex
Report section on experiments for standard simulation

\documentclass[twoside]{article}

\usepackage{lipsum} % Package to generate dummy text throughout this template

@@

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage[hidelinks]{hyperref} % For hyperlinks in the PDF

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text
\usepackage{paralist} % Used for the compactitem environment which makes bullet points with less space between them

\usepackage{todonotes}
\usepackage{amsmath,amssymb,amsfonts}
\providecommand{\abs}[1]{\left \lvert #1 \right \rvert}

@@

\begin{multicols}{2} % Two-column layout throughout the main article text

\section{Introduction}
\lettrine[nindent=0em,lines=3]{S} imulations of large hyperelastic materials
can be very time consuming. When working with materials of varying density
of vertices/scale of mesh, the highest densities often give the most
strict bounds on size of the time step of the simulation in order to keep
the materials (and grids) stable and the discretization correct. In a
``standard'' simulation with a global time step size we need to compute
possibly unnecessarily many computations on the coarser parts of the material
caused by this bound. When furthermore having materials with a large amount
of somewhat equally distributed vertices and smaller patches of detailed
areas with higher density of vertices, it would be an advantage to be able
to perform large time steps for the majority of the material (the coarse
parts) and smaller steps for the local patches where vertices could cause
instability. We call the concept of varying the timestep local adaptive
optimization of time step (LAOTS), in which ``local'' either refers to spatial
locality (as described above) or time locality where the time step is global
for all vertices but varies during the simulation. We will in this report
study possible approaches for application of LAOTS. In order to keep the focus
on possible LAOTS methods and remove unneccessary complexities, we will study
the simple 1 dimensional simulation of a mass-spring system and try to apply
LAOTS for it. Afterwards we will shortly discuss the application of our gained
knowledge to the domain of hyperelastic material simulation and simulations in
general.

@@

\section{Methods}

@@

When simulating the mass-spring system we perform the following steps:
\begin{compactenum} \item Initialize system \item Calculate forces \item
Calculate derrivatives \item Update positions and velocities \item Update time
\item Go to step 2 if the simulation has not reached it's maximum simulation
time. \end{compactenum} We now describe the details of each of the steps of
the simulation.

@@

The total force of a spring is the sum of the spring force $F_{h,d}$
calculated by using Hooke's Law and the damping force $F_{C,d}$
using a simple damping method [TODO: Add reference (wiki):
\url{http://en.wikipedia.org/wiki/Harmonic_oscillator}]: \begin{align}
\nonumber l_d &= p_{l,d} - p_{r,d} \\ \label{eq:hooke} F_{h,d} &= -k_d x_d
= -k_d (l_d - l_{0,d}) \\ \label{eq:damping} F_{C,d} &= -C_d \frac{\partial
x_d}{ \partial t} = - C_d \frac{\left(v_{l,d} - v_{r,d}\right) l_d}{\abs{l_d}}
\\ F_d &= F_{h,d} + F_{C,d} \end{align} [TODO: Maybe add direction correction:
$F_d \cdot (-l_d / \abs{l_d})$] in which $p_{l,d}$ and $p_{r,d}$ are the
positions of the left and right endpoints of the spring $d$ respectively, and
$v_{l,d}$ and $v_{r,d}$ are the velocities of the left and right endpoint
vertices of the spring $d$ respectively, and $F_d$ is the total force of
spring $d$.

@@

We now know how to calculate the force of the spring system and apply it
to the vertices. In order to calculate the derrivatives of the system
(step 3) we use Newton's second law of motion [TODO: Add reference,
\url{http://en.wikipedia.org/wiki/Newton's_laws_of_motion}] to get the
difference in speed $\Delta u_v$ and position $\Delta p_v$ of each vertex $v$
using a given time step size $\Delta t$: \begin{align} \label{eq:delta_u}
\Delta u_v &= \frac{\partial u_v}{\partial t} \Delta t = \frac{F_v}{m_v}
\Delta t \\ \label{eq:delta_p} \Delta p_v &= u \Delta t \end{align}

As mentioned in the beginning of this section we have chosen to use the
explicit Euler method defined below in equation \ref{eq:explicit_euler}
to calculate the time ``integration'' [TODO: Add reference,
\url{http://en.wikipedia.org/wiki/Euler_method}] when stepping through
the simulation. Given a variable $x$ as a function of the time $t$, the
variable $x$ at the time $t + \Delta t$ is defined as: \begin{align}
\label{eq:explicit_euler} x(t+\Delta t) &= x(t) + \Delta t \frac{\partial
x(t)}{\partial t} \end{align} Applying this method for the timestepping
of our simulation we get the two following position and velocity update
equations using equations \ref{eq:delta_u} and \ref{eq:delta_p}: \begin{align}
p_v(t+\Delta t) &= p_v(t) + \Delta p_v \\ u_v(t+\Delta t) &= u_v(t) + \Delta
u_v \end{align} We use these equations to perform step 4 of the simulation.

@@

\subsection{Atomic updating using heuristics}
The second LAOT method we propose is space-adaptive. We perform simulations
using a fixed timestep size. In each iteration we perform position update of
all vertices, but only re-calculate acceleration and velocity of the springs
and points that fulfill one of the defined heuristics. The heuristics can be
chosen to fit the domain of the simulation, but we have chosen to use the
number of steps since last update and the ratio between the current spring
length and the spring length at the last update. This should force a bound on
the maximal timestep before update as well as making sure updates are done
more frequently in case the springlength is changing more rapidly. Further
details are explained in section \ref{sec:experiments}.

@@

Show results for:

\begin{itemize}
    \item Normal simulation
    \item Switch
    \item Atomic
    \item Inverse
\end{itemize}

Examples of

%\begin{table}[H]
%\caption{Example table}
%\centering
%\begin{tabular}{llr}
%\toprule
%\multicolumn{2}{c}{Name} \\
%\cmidrule(r){1-2}
%First name & Last Name & Grade \\
%\midrule
%John & Doe & $7.5$ \\
%Richard & Miles & $2$ \\
%\bottomrule
%\end{tabular}
%\end{table}

@@

\end{multicols}

>>>>

\documentclass[11pt]{article}

@@

%\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
%\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage[hidelinks]{hyperref} % For hyperlinks in the PDF

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text
%\usepackage{paralist} % Used for the compactitem environment which makes bullet points with less space between them

\usepackage{todonotes}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{subcaption}
\providecommand{\abs}[1]{\left \lvert #1 \right \rvert}

@@

%\begin{multicols}{2} % Two-column layout throughout the main article text

\section{Introduction}
\lettrine[nindent=0em,lines=3]{S} imulations of large hyperelastic materials
can be very time consuming. When working with materials of varying density
of vertices/scale of mesh, the highest densities often give the most
strict bounds on size of the time step of the simulation in order to keep
the materials (and grids) stable and the discretization correct.

In a ``standard'' simulation with a global time step size we need to compute
possibly unnecessarily many computations on the coarser parts of the material
caused by this bound. When furthermore having materials with a large amount
of somewhat equally distributed vertices and smaller patches of detailed
areas with higher density of vertices, it would be an advantage to be able
to perform large time steps for the majority of the material (the coarse
parts) and smaller steps for the local patches where vertices could cause
instability. We call the concept of varying the timestep local adaptive
optimization of time step (LAOTS), in which ``local'' either refers to spatial
locality (as described above) or time locality where the time step is global
for all vertices but varies during the simulation.

We will in this report
study possible approaches for application of LAOTS. In order to keep the focus
on possible LAOTS methods and remove unneccessary complexities, we will study
the simple 1 dimensional simulation of a mass-spring system and try to apply
LAOTS for it. Afterwards we will shortly discuss the application of our gained
knowledge to the domain of hyperelastic material simulation and simulations in
general.

@@

\section{Methods}
\label{sec:methods}

@@

When simulating the mass-spring system we perform the following steps:
\begin{enumerate} \item Initialize system \item Calculate forces \item
Calculate derrivatives \item Update positions and velocities \item Update time
\item Go to step 2 if the simulation has not reached it's maximum simulation
time. \end{enumerate} We now describe the details of each of the steps of
the simulation.

@@

The total force of a spring is the sum of the spring force $F_{h,d}$
calculated by using Hooke's Law and the damping force $F_{C,d}$
using a simple damping method \todo{Add reference (wiki):
    \url{http://en.wikipedia.org/wiki/Harmonic_oscillator}}: \begin{align}
\nonumber l_d &= p_{l,d} - p_{r,d} \\ \label{eq:hooke} F_{h,d} &= -k_d x_d
= -k_d (l_d - l_{0,d}) \\ \label{eq:damping} F_{C,d} &= -C_d \frac{\partial
x_d}{ \partial t} = - C_d \frac{\left(v_{l,d} - v_{r,d}\right) l_d}{\abs{l_d}}
\\ F_d &= F_{h,d} + F_{C,d} \end{align} \todo{Maybe add direction correction:
    $F_d \cdot (-l_d / \abs{l_d})$} in which $p_{l,d}$ and $p_{r,d}$ are the
positions of the left and right endpoints of the spring $d$ respectively, and
$v_{l,d}$ and $v_{r,d}$ are the velocities of the left and right endpoint
vertices of the spring $d$ respectively, and $F_d$ is the total force of
spring $d$.

@@

We now know how to calculate the force of the spring system and apply it
to the vertices. In order to calculate the derrivatives of the system
(step 3) we use Newton's second law of motion \todo{Add reference,
    \url{http://en.wikipedia.org/wiki/Newton's_laws_of_motion}} to get the
difference in speed $\Delta u_v$ and position $\Delta p_v$ of each vertex $v$
using a given time step size $\Delta t$: \begin{align} \label{eq:delta_u}
\Delta u_v &= \frac{\partial u_v}{\partial t} \Delta t = \frac{F_v}{m_v}
\Delta t \\ \label{eq:delta_p} \Delta p_v &= u \Delta t \end{align}

As mentioned in the beginning of this section we have chosen to use the
explicit Euler method defined below in equation \ref{eq:explicit_euler}
to calculate the time ``integration'' \todo{Add reference,
    \url{http://en.wikipedia.org/wiki/Euler_method}} when stepping through
the simulation. Given a variable $x$ as a function of the time $t$, the
variable $x$ at the time $t + \Delta t$ is defined as: \begin{align}
\label{eq:explicit_euler} x(t+\Delta t) &= x(t) + \Delta t \frac{\partial
x(t)}{\partial t} \end{align} Applying this method for the timestepping
of our simulation we get the two following position and velocity update
equations using equations \ref{eq:delta_u} and \ref{eq:delta_p}: \begin{align}
p_v(t+\Delta t) &= p_v(t) + \Delta p_v \\ u_v(t+\Delta t) &= u_v(t) + \Delta
u_v \end{align} We use these equations to perform step 4 of the simulation.

@@

\subsection{Atomic updating using heuristics}
The second LAOT method we propose is space-adaptive. We perform simulations
using a fixed timestep size. In each iteration we perform position update of
all vertices, but only re-calculate acceleration and velocity of the springs
and points that fulfill some defined heuristics.

The heuristics can be chosen to fit the domain of the simulation, but we have
chosen to use the number of steps since last update and the ratio between the
current spring length and the spring length at the last update. This should
force a bound on the maximal timestep before update as well as making sure
updates are done more frequently in case the springlength is changing more
rapidly. Further details are explained in section \ref{sec:experiments}.

@@

We first give an example of the actual application of LAOTS using the standard
simulation of mass spring particles. Figure \ref{fig:std} shows two simulations
of the same set of particles distributed with two different scales simulated for
500 seconds each. All the springs $s \in S$ are initiated in a stretched state/position
with a length 1.25 times longer than the rest length, damping and stiffness
constants $C_s = 0.5, K_s = 0.5$, and all vertices $v \in V$
have a mass of $m_v = 30$.
\ref{fig:std_multi_30fps} shows a simulation using 30 fps and Figure
\ref{fig:std_multi_1fps} shows a simulation using 1 fps. We clearly see that
using a timestep of 30 fps the simulation is stable and able to handle the
compression of all the small springs around $t = 110$ whereas a switch occurs
when simulating at 1 fps. This switch causes ripples of sideeffects for the rest
of the simulation and hence the final result is incorrect. No switches of the longer springs are however occuring and hence these pose no problem at 1 fps. We will use this simulation for
testing all three LAOTS methods proposed in section \ref{sec:methods}.

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_multiscale_30fps.png}
        \caption{30 fps}
        \label{fig:std_multi_30fps}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_multiscale_1fps.png}
        \caption{1 fps}
        \label{fig:std_multi_1fps}
    \end{subfigure}
    \caption{Standard simulations of particles distributed with two scales using
    different timestep sizes}
    \label{fig:std}
\end{figure}

\todo{Write something about the explicit euler based on uniform}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_uniform_30fps.png}
        \caption{30 fps}
        \label{fig:std_uni_30fps}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{../images/standard_uniform_1fps.png}
        \caption{1 fps}
        \label{fig:std_uni_1fps}
    \end{subfigure}
    \caption{Standard simulations of particles distributed uniformly using
    different timestep sizes}
\end{figure}

\subsection{Adapt-on-switch}
Special things to note: How subdivision is done: new fixed (30) or relative (dt / 10)
new fixed: We don't catch it early enough. Maybe something with the explicit
euler, maybe something with the linear model

Relative: We get too fine-grained - maybe related to the above

\subsection{Atomic updating using heuristics}

Heuristic: steps since last update and ratio of length since last update

\subsection{Global estimated acceleration}

%Examples of
%
%\begin{table}[H]
%\caption{Example table}
%\centering
%\begin{tabular}{llr}
%\toprule
%\multicolumn{2}{c}{Name} \\
%\cmidrule(r){1-2}
%First name & Last Name & Grade \\
%\midrule
%John & Doe & $7.5$ \\
%Richard & Miles & $2$ \\
%\bottomrule
%\end{tabular}
%\end{table}

@@
